---
title: "Likelihood"
description: |
  Material covered on Day 2.
author: 
  - name: Dominique Gravel
  - name: Andrew MacDonald
date: 2021-08-16
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_float: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)


make_url <- function(pres_name){
  paste0("https://bios2.github.io/BiodiversityModelling/02_likelihood/",
         pres_name,
         "/#1")
}


embed_slides <- function(pres_name){
  
  url <- paste0("https://bios2.github.io/BiodiversityModelling/02_likelihood/",
                pres_name,
                "/#1")
  
  iframe <- xaringanExtra::embed_xaringan(url)
  
  lnk <- paste0("[Click to view full screen](", url, ")")
  
  knitr::asis_output(paste0(iframe,"\n", lnk))

}

slide_link <- function(pres_name){
  lnk <- paste0("[Click to view full screen](", make_url(pres_name), ")")
  knitr::asis_output(lnk)
}
```

## Introduction to probability 

```{r}
xaringanExtra::embed_xaringan(make_url("01-likelihood"))
slide_link("01-likelihood")
```

## Sutton trees

```{r}
maple <- readr::read_csv2("https://raw.githubusercontent.com/bios2/bios2.github.io/master/_BiodiversityModelling/02_likelihood/01-likelihood/data/sutton.csv")

```


## Statistical field work -- exploring your dataset

**Step 1.** Return to your hypothesized distribution. Find its random number generation function (e.g. `rbinom()` for the Binomial Distribution). Experiment with different parameters values to create data that looks close to your own data.

**Step 2.** Attempt to fit the same parameters by Maximum likelihood. If you distribution has two parameters, fix one and find the Maximum Likelihood value of the other. **Stretch goal** Try to fit both of them at once.

**Step 3.** Look at the help for `?MASS::fitdistr()`. Try to use it to find your distribution parameters. 

### Continuous distributions

| distribution | positive or negative?       | Quick description                                              |
|--------------|-----------------------------|----------------------------------------------------------------|
| Normal       | positive or negative values | Can result from many small effects added together              |
| Lornormal    | positive values only        | results from many things multiplied together                   |
| Gamma        | positive values only        | can be the time to wait for a given number of things to happen |
| Exponential  | Positive only               | Lengths of time between random events                          |
| Beta         | Positive only               | Between 0 and 1. Can be any proportion.                        |

### Discrete distributions


| distribution      | positive or negative? | Quick description                                                                                                     |
|-------------------|-----------------------|-----------------------------------------------------------------------------------------------------------------------|
| Poisson           | Positive              | Counting things that occur randomly over time                                                                         |
| Negative Binomial | Positive              | Counts of things as in the Poisson, but more variable. Also: the number of trials till a certain number of successes. |
| Binomial          | Positive              | Number of "successes" out of a number of trials, when probability of success is always the same                       |
