---
title: "Intro to Bayesian Statistics"
description: |
  Material covered on Day 4.
author: 
  - name: Dominique Gravel
  - name: Andrew MacDonald
date: 2021-08-28
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_float: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

```

## Bayes slides, Tuesday

```{r}
xaringanExtra::embed_xaringan("https://bios2.github.io/biodivmodelling2021/Bayes.html#1")
```

## Wednesday Problem

```{r}
xaringanExtra::embed_xaringan("https://bios2.github.io/biodivmodelling2021/tree_models.html#1")
```


## Generative (multilevel) models 

```{r}
xaringanExtra::embed_xaringan("https://bios2.github.io/biodivmodelling2021/generative.html#1")
```



## Field data project

Try to find a conjugate prior for part of your data. If you have a two-parameter distribution (e.g. the Normal distribution) you may only be able to use a conjugate prior for one parameter. 

Estimate your distribution parameters with this approach. Do you get a similar response to previously? 


## Mixed effect model

$$
\begin{align}
y &\sim \text{Normal}(\mu, \sigma_y) \\
\mu &= \beta_0 + \beta_1 x + \alpha_{block} \\
\alpha_{block} &\sim \text{Normal}(0, \sigma_{block}) \\
\sigma_y &\sim \text{Exponential}(2) \\
\sigma_{block} &\sim  \text{Exponential}(4) \\
\beta_0 &\sim \text{Normal}(7, 0.2) \\
\beta_1 &\sim \text{Normal}(4, 0.5)
\end{align}
$$
